---
title: "Biodiversity_Riparian_Forest"
author: "Camila Rodriguez"
date: "2025-03-05"
output: html_document
---

```{r setup, include=FALSE}

Fish_salamander<- read.csv('Data/AS01001_v1.csv')
invertebrate<- read.csv('Data/AS01002_v1.csv')
transect_data_sites<- read.csv('Data/AS01003_v1.csv')
wood_measurements<- read.csv('Data/AS01004_v1.csv')
pool_measurement<-read.csv('Data/AS01005_v1.csv')

```

```{r load packages, message=FALSE, warning=FALSE}
library(ggplot2)
library (patchwork)
library(dplyr)
```

```{r figure 1, fig.cap = "Distribution of stream characteristics (maximum depth, depth at outflow, stream width, and stream length) in old-growth forest riparian zones and secondary growth riparian zones.}
binnumber <- 2 * (172)^(1/3)
binnumber

depth_plot <-  ggplot(pool_measurement, aes(x = DEPTH_MAX, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Depth (cm)",
y = "Count", title = "Figure 1. Distribution of stream characteristics by forest treatment") +
   theme_bw() + theme(legend.position = "none") 

flowdepth_plot <-  ggplot(pool_measurement, aes(x = DEPTH_OUTFLOW, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Depth at stream outflow (cm)",
y = "Count") +
  theme_bw() + theme(legend.position = "none") 

width_plot <-  ggplot(pool_measurement, aes(x = WIDTH, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Pool width (cm)",
y = "Count") +
   theme_bw() + theme(legend.position = "none")

length_plot <-  ggplot(pool_measurement, aes(x = POOL_LENGTH, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Pool length (cm)",
y = "Count", caption = "Figure 1. Distribution of stream characteristics
(maximum depth, depth at outflow, stream width, and stream length) in 
old-growth forest riparian zones and secondary growth riparian zones. 
Data collected July-October 2014 by HJ Andrews LTER") +
  theme_bw() + theme(legend.position = "none")


stream_qualities_plot <- depth_plot / flowdepth_plot / width_plot / length_plot 
stream_qualities_plot

ggsave("stream_qualities_plot.jpeg", stream_qualities_plot, width = 8, height = 14)

```

```{r other-plots}
(hist.maxdepth <- ggplot(pool_measurement, aes(x=DEPTH_MAX))+ geom_histogram(bins = 11) + theme_bw())
(hist.outflowdepth <- ggplot(pool_measurement, aes(x=DEPTH_OUTFLOW))+ geom_histogram(bins = 11) + theme_bw())
(hist.poollength <- ggplot(pool_measurement, aes(x=POOL_LENGTH))+ geom_histogram(bins = 11) + theme_bw())
(hist.width <- ggplot(pool_measurement, aes(x=WIDTH))+ geom_histogram(bins = 11) + theme_bw())
```

```{r figure2, fig.cap = "Biodiversity Patterns across Forest Types: Species Richness, Species Abundance, and Shannon Diversity Index.", message=FALSE, warning=FALSE}

# Calculate species richness: count of unique species per forest type
species_richness <- Fish_salamander %>%
  group_by(REACH_TYPE) %>%
  summarise(Richness = n_distinct(SPECIES))

# Calculate species abundance: total count of individuals per forest type
species_abundance <- Fish_salamander %>%
  group_by(REACH_TYPE) %>%
  summarise(Abundance = n())

# Calculate Shannon diversity index:
# First, calculate counts per species for each forest type, then compute the index.
shannon_data <- Fish_salamander %>%
  group_by(REACH_TYPE, SPECIES) %>%
  summarise(Count = n()) %>%
  group_by(REACH_TYPE) %>%
  summarise(Diversity = -sum((Count / sum(Count)) * log(Count / sum(Count))))

# Create a bar plot for species richness
richness_plot <- ggplot(species_richness, aes(x = REACH_TYPE, y = Richness, fill = REACH_TYPE)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("seagreen3", "dodgerblue")) +
  labs(x = "Forest Type", y = "Unique Species Count", title = "Species Richness by Forest Type") +
  theme_bw() + theme(legend.position = "none")

# Create a bar plot for species abundance
abundance_plot <- ggplot(species_abundance, aes(x = REACH_TYPE, y = Abundance, fill = REACH_TYPE)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("seagreen3", "dodgerblue")) +
  labs(x = "Forest Type", y = "Total Individual Count", title = "Species Abundance by Forest Type") +
  theme_bw() + theme(legend.position = "none")

# Create a bar plot for the Shannon diversity index
diversity_plot <- ggplot(shannon_data, aes(x = REACH_TYPE, y = Diversity, fill = REACH_TYPE)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("seagreen3", "dodgerblue")) +
  labs(x = "Forest Type", y = "Shannon Diversity Index", title = "Species Diversity by Forest Type") +
  theme_bw() + theme(legend.position = "none")

# Combine the three plots vertically using patchwork
biodiversity_plots <- richness_plot / abundance_plot / diversity_plot

# Save the combined figure to a JPEG file
ggsave("biodiversity_patterns.jpeg", biodiversity_plots, width = 8, height = 12)

# Display the combined figure
biodiversity_plots