---
title: "Biodiversity_Riparian_Forest"
author: "Camila Rodriguez"
date: "2025-03-05"
output: html_document
---
## Setup and load datasets 
```{r setup, include=FALSE}

Fish_salamander<- read.csv('Data/AS01001_v1.csv')
#invertebrate<- read.csv('Data/AS01002_v1.csv') provided but not needed for our analysis
stream_data<- read.csv('Data/AS01003_v1.csv')
wood_measurements<- read.csv('Data/AS01004_v1.csv')
pool_measurement<-read.csv('Data/AS01005_v1.csv')

#install.packages("dplyr")
#install.packages("tidyverse")
library(ggplot2)
library (patchwork)
library(dplyr)
library(tidyverse)
library(here)
library(lubridate)
library(nlme)
library(gtsummary)
library(lme4)

```

### Exploratory Image 1 
```{r figure 1, fig.cap = "Distribution of stream characteristics (maximum depth, depth at outflow, stream width, and stream length) in old-growth forest riparian zones and secondary growth riparian zones.}
binnumber <- 2 * (172)^(1/3)
binnumber

depth_plot <-  ggplot(pool_measurement, aes(x = DEPTH_MAX, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Depth (cm)",
y = "Count", title = "Figure 1. Distribution of stream characteristics by forest treatment") +
   theme_bw() + theme(legend.position = "none") 

flowdepth_plot <-  ggplot(pool_measurement, aes(x = DEPTH_OUTFLOW, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Depth at stream outflow (cm)",
y = "Count") +
  theme_bw() + theme(legend.position = "none") 

width_plot <-  ggplot(pool_measurement, aes(x = WIDTH, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Pool width (cm)",
y = "Count") +
   theme_bw() + theme(legend.position = "none")

length_plot <-  ggplot(pool_measurement, aes(x = POOL_LENGTH, fill = REACH_TYPE)) +
geom_histogram(bins = 11) + # creates histograms
  facet_wrap (~ REACH_TYPE) +
  scale_fill_manual(values= c('seagreen3', 'dodgerblue')) +
labs(x = "Pool length (cm)",
y = "Count", caption = "Figure 1. Distribution of stream characteristics
(maximum depth, depth at outflow, stream width, and stream length) in 
old-growth forest riparian zones and secondary growth riparian zones. 
Data collected July-October 2014 by HJ Andrews LTER") +
  theme_bw() + theme(legend.position = "none")


stream_qualities_plot <- depth_plot / flowdepth_plot / width_plot / length_plot 
stream_qualities_plot

ggsave("stream_qualities_plot.jpeg", stream_qualities_plot, width = 8, height = 14)

```

### Exploratory histograms 
```{r other-plots}
(hist.maxdepth <- ggplot(pool_measurement, aes(x=DEPTH_MAX))+ geom_histogram(bins = 11) + theme_bw())
(hist.outflowdepth <- ggplot(pool_measurement, aes(x=DEPTH_OUTFLOW))+ geom_histogram(bins = 11) + theme_bw())
(hist.poollength <- ggplot(pool_measurement, aes(x=POOL_LENGTH))+ geom_histogram(bins = 11) + theme_bw())
(hist.width <- ggplot(pool_measurement, aes(x=WIDTH))+ geom_histogram(bins = 11) + theme_bw())
##

```

### Exploratory Image 2
```{r figure2, fig.cap = "Canopy Cover vs. Chlorophyll-a and Woody Debris Effect on Stream Structure"}
# Canopy Cover vs. Chlorophyll-a (LOCKED)
# Based on visible clues in the data, we removed points where canopy cover is < 35%.
clean_canopy_data <- stream_data %>%
  filter(CANOPY_COVER >= 35)

canopy_plot <- ggplot(clean_canopy_data, aes(x = REACH_TYPE, y = CHL_A_TILES, fill = REACH_TYPE)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(aes(color = REACH_TYPE), width = 0.2, size = 1.5, alpha = 0.5) +
  scale_fill_manual(values = c("seagreen3", "dodgerblue")) +
  scale_color_manual(values = c("seagreen4", "blue4")) +
  labs(
    x = "Forest Type",
    y = "Chlorophyll-a Concentration (µg/cm²)",
    title = "Chlorophyll-a Levels in Old-Growth vs. Secondary Forests"
  ) +
  theme_bw() +
  theme(legend.position = "none")

# Separate Histograms for OG and SG
wood_histogram <- ggplot(wood_measurements, aes(x = WOOD_LENGTH, fill = REACH_TYPE)) +
  geom_histogram(bins = 15, alpha = 0.7) +  # Individual histograms
  scale_fill_manual(values = c("seagreen3", "dodgerblue")) +
  scale_x_log10() +  # Log scale for better spread
  facet_wrap(~ REACH_TYPE) +  # Separate plots for OG and SG
  labs(
    x = "Woody Debris Length (log scale, cm)",
    y = "Count",
    title = "Woody Debris Size Distribution by Forest Type"
  ) +
  theme_bw() +
  theme(legend.position = "none")

# Combine Both Plots into One Figure with Proper Caption Spacing 
combined_figure <- canopy_plot / wood_histogram +
  plot_annotation(
    caption = "Figure 2. Distribution of woody debris sizes in old-growth forest riparian zones and secondary growth riparian zones. Data collected July-October 2014 by HJ Andrews LTER."
  ) &
  theme(
    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic")  # Ensures proper alignment
  )

# Save the figure with increased width
ggsave("canopy_woodydebris.jpeg", combined_figure, width = 14, height = 10)  # Increased width

# Display the figure
combined_figure
```

## Abundance of salamanders 
```{r}

salamander<- Fish_salamander %>% filter(SPECIES== "DITE")%>%
  mutate(SPECIES_NAME = "Dicamptodon tenebrosus (Coastal Giant Salamander)")

salamander_abundance_by_pair <- function(salamander, correction_factor = 1.34) {
  
  salamander <- salamander %>%
    mutate(SURVEY_METHOD = toupper(SURVEY_METHOD),
           MARKED_UNMARKED = toupper(MARKED_UNMARKED))
  
  # Combine SITE and REACH_TYPE to define unique pairs
  site_pairs <- salamander %>%
    select(SITE, REACH_TYPE) %>%
    distinct()
  
  results <- data.frame(
    SITE = character(),
    REACH_TYPE = character(),
    Abundance_Estimate = numeric(),
    Method = character(),
    stringsAsFactors = FALSE
  )
  
  for (i in 1:nrow(site_pairs)) {
    site <- site_pairs$SITE[i]
    reach <- site_pairs$REACH_TYPE[i]
    
    pair_data <- salamander %>%
      filter(SITE == site, REACH_TYPE == reach)
    
    if (site %in% c("MCT_E", "MCT_W")) {
      raw_count <- pair_data %>%
        filter(SURVEY_METHOD %in% c("P1", "P2", "P3")) %>%
        nrow()
      est <- raw_count * correction_factor
      method <- "Depletion (corrected)"
      
    } else if (all(c("M", "R") %in% unique(pair_data$SURVEY_METHOD))) {
      M <- pair_data %>%
        filter(SURVEY_METHOD == "M", MARKED_UNMARKED == "M") %>%
        nrow()
      C <- pair_data %>%
        filter(SURVEY_METHOD == "R") %>%
        nrow()
      R <- pair_data %>%
        filter(SURVEY_METHOD == "R", MARKED_UNMARKED == "M") %>%
        nrow()
      
      est <- if (R > 0) ((M + 1) * (C + 1)) / (R + 1) - 1 else NA
      method <- "Mark-recapture"
      
    } else {
      est <- NA
      method <- "Insufficient data"
    }
    
    results <- rbind(
      results,
      data.frame(SITE = site, REACH_TYPE = reach,
                 Abundance_Estimate = round(est, 2),
                 Method = method,
                 stringsAsFactors = FALSE)
    )
  }
  
  return(results)
}

abundance_by_pair <- salamander_abundance_by_pair(salamander)
print(abundance_by_pair)

## We need 18 sites (9 paired sites)- need to adjust code to include 3 msising sites that were filtered for NA Data?

```
### Join Tables for Regression - not using
```{r}
full_salamander_data_old <- abundance_by_pair %>%
  left_join(stream_data, by = c("SITE", "REACH_TYPE")) %>%
  left_join(pool_measurement, by = c("SITE", "REACH_TYPE")) %>%
  left_join(wood_measurements, by = c("SITE", "REACH_TYPE"))%>%
  filter(!is.na(Abundance_Estimate)) %>%
  drop_na()
``` 
### Joining tables by aggregating stream, wood and pool for site and reach type
```{r}
# 1. Summarize stream data: includes width, canopy, and chlorophyll
stream_summary <- stream_data %>%
  group_by(SITE, REACH_TYPE) %>%
  summarise(
    BANKFULL_WIDTH = mean(BANKFULL_WIDTH, na.rm = TRUE),
    WETTED_WIDTH = mean(WETTED_WIDTH, na.rm = TRUE),
    CANOPY_COVER = mean(CANOPY_COVER, na.rm = TRUE),
    CHL_A_TILES = mean(CHL_A_TILES, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Summarize pool measurements: depth and pool structure
pool_summary <- pool_measurement %>%
  group_by(SITE, REACH_TYPE) %>%
  summarise(
    DEPTH_MAX = mean(DEPTH_MAX, na.rm = TRUE),
    DEPTH_OUTFLOW = mean(DEPTH_OUTFLOW, na.rm = TRUE),
    DEPTH_RESIDUAL = mean(DEPTH_RESIDUAL, na.rm = TRUE),
    POOL_LENGTH = mean(POOL_LENGTH, na.rm = TRUE),
    WIDTH = mean(WIDTH, na.rm = TRUE),
    .groups = "drop"
  )

# 3. Summarize wood measurements: size of woody debris
wood_summary <- wood_measurements %>%
  group_by(SITE, REACH_TYPE) %>%
  summarise(
    WOOD_LENGTH = mean(WOOD_LENGTH, na.rm = TRUE),
    DIAMETER_1 = mean(DIAMETER_1, na.rm = TRUE),
    DIAMETER_2 = mean(DIAMETER_2, na.rm = TRUE),
    .groups = "drop"
  )

# 4. Join all summaries with abundance estimates
full_salamander_data <- abundance_by_pair %>%
  left_join(stream_summary, by = c("SITE", "REACH_TYPE")) %>%
  left_join(pool_summary, by = c("SITE", "REACH_TYPE")) %>%
  left_join(wood_summary, by = c("SITE", "REACH_TYPE")) %>%
  filter(!is.na(Abundance_Estimate)) %>%
  drop_na()

```

### Histogram of Abundance Estimates
```{r}
ggplot(full_salamander_data, aes(x = Abundance_Estimate)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, 
                 fill = "lightblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(
    title = "Distribution of Salamander Abundance Estimates (One Row per Site)",
    x = "Abundance Estimate",
    y = "Density"
  ) +
  theme_bw()

# Shapiro-Wilk normality test (valid if sample size is between 3 and 5000)
shapiro.test(full_salamander_data$Abundance_Estimate)
```

### Log Transformation
```{r}
# Add a small constant in case there are zeroes (shouldn't be any, but just in case)
full_salamander_data$log_abundance <- log(full_salamander_data$Abundance_Estimate)

# Check distribution again
ggplot(full_salamander_data, aes(x = log_abundance)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Log-Transformed Salamander Abundance", x = "log(Abundance)", y = "Density") +
  theme_bw()

# Shapiro-Wilk test on log-transformed values
shapiro.test(full_salamander_data$log_abundance)
```


### Model 1
```{r}
#Salamander abundance as a function of canopy cover and wetted width (poisson)
#DITE – Dicamptodon tenebrosus (Coastal Giant Salamander)
#Poisson regression because we are interested in abundance 

model1 <- lm(log_abundance ~ CANOPY_COVER + WETTED_WIDTH, data = full_salamander_data)
summary(model1)
plot(model1)
AIC(model1)

```
### Model 2
```{r}
#Salamander abundance as a function of downed woody debris length and max pool depth
# added wood length as a funcction of reach type because larger pieces of debris are found in old growth 

#residuals have a pattern
model2 <- lm(log_abundance ~ WOOD_LENGTH*REACH_TYPE + DEPTH_MAX*REACH_TYPE + factor(SITE), data = full_salamander_data)
summary(model2)
plot(model2)
AIC(model2)

#removing factor makes Q-Q more linear
model2B <- lm(log_abundance ~ WOOD_LENGTH*REACH_TYPE + DEPTH_MAX*REACH_TYPE, data = full_salamander_data)
summary(model2B)
plot(model2B)
AIC(model2B)

#removing outlier site
full_salamander_data_no_COOK <- full_salamander_data[full_salamander_data$SITE != "COOK", ]

model2C <- lm(log_abundance ~ WOOD_LENGTH*REACH_TYPE + DEPTH_MAX*REACH_TYPE, data = full_salamander_data_no_COOK )
summary(model2C)
plot(model2C)
AIC(model2C)

model2D <- lm(log_abundance ~ WOOD_LENGTH*REACH_TYPE + DEPTH_MAX*REACH_TYPE + factor(SITE), data = full_salamander_data_no_COOK)
summary(model2D)
plot(model2D)
AIC(model2D)
```
<<<<<<< HEAD
```{r VT experiment with data frame}
salamander_only <- Fish_salamander %>% filter(SPECIES== "DITE")%>%
  mutate(SPECIES_NAME = "Dicamptodon tenebrosus (Coastal Giant Salamander)")

salamander_only_basic <- salamander_only %>% select(-c(DBCODE, ENTITY, DATE, MARKED_UNMARKED, VERT_NO, SPECIES, SPP_LIFESTAGE, TAIL_LENGTH, WEIGHT, SPECIES_NAME))

salamander_only_basic_reordered <- salamander_only_basic[, c("REACH_TYPE", "SURVEY_METHOD", "SITE", "VENT_LENGTH")] 

salamander_condensed <- salamander_only_basic_reordered %>%
  group_by(SITE,REACH_TYPE, SURVEY_METHOD) %>%
  summarize(avg_length = mean(VENT_LENGTH, na.rm = TRUE), .groups = "drop")

wood_basic <- wood_measurements %>% select(-c(DBCODE, ENTITY, DATE, DIAMETER_1, DIAMETER_2, WOOD_LENGTH))
wood_condensed <- wood_basic %>%
  group_by(SITE,REACH_TYPE) %>%
  summarize(number_of_wood_pieces = max(WOOD_PIECE, na.rm = TRUE), .groups = "drop")

salamanders_and_wood <- salamander_condensed  %>%
  left_join(wood_condensed, by = c("SITE", "REACH_TYPE"))

water_basic <- pool_measurement %>% select(-c(DBCODE, ENTITY, DATE, POOL_NO, DEPTH_OUTFLOW, DEPTH_RESIDUAL, POOL_LENGTH, WIDTH))

water_condensed <- water_basic %>%
  group_by(SITE,REACH_TYPE) %>%
  summarize(average_maximum_depth= mean(DEPTH_MAX, na.rm = TRUE), .groups = "drop")

salamanders_and_wood_and_water <- salamanders_and_wood  %>%
  left_join(water_condensed , by = c("SITE", "REACH_TYPE"))

stream_basic <- stream_data %>% select(-c(DBCODE, ENTITY, DATE, METER, BANKFULL_WIDTH, WETTED_WIDTH, DATE_DEPLOYED_TILES, DATE_RETRIEVED_TILES, CHL_A_TILES))

stream_condensed <- stream_basic %>%
  group_by(SITE,REACH_TYPE) %>%
  summarize(average_canopy_cover= mean(CANOPY_COVER, na.rm = TRUE), .groups = "drop")


salamanders_and_wood_and_water_and_stream <- salamanders_and_wood_and_water  %>%
  left_join(stream_condensed  , by = c("SITE", "REACH_TYPE"))

#make all categorical variables factors 
str(salamanders_and_wood_and_water_and_stream)

salamanders_and_wood_and_water_and_stream <- salamanders_and_wood_and_water_and_stream %>% 
  mutate(SITE= factor(SITE, levels = c("COOK", "FRITZ", "LO701", "LO703", "MCT_E", "MCT_W", "MR404", "STR")),
         REACH_TYPE = factor(REACH_TYPE, levels= c("SG", "OG")),
         SURVEY_METHOD = factor(SURVEY_METHOD, levels = c("M", "R", "P1", "P2", "P3")))

#fit linear regression
sala.lm1 <- lm(avg_length ~ SITE + REACH_TYPE + number_of_wood_pieces, data = salamanders_and_wood_and_water_and_stream)
summary(sala.lm1)
plot(sala.lm1)

#fit linear regression with random effect 
sala.lm2 <- gls(avg_length ~ SITE + REACH_TYPE + number_of_wood_pieces, data = salamanders_and_wood_and_water_and_stream)

#for a multi-level model with random intercept by reach nested in plot

#wood pieces
sala.lmm3 <- lme(avg_length ~ SITE + REACH_TYPE + number_of_wood_pieces, random =~1 | REACH_TYPE/SITE, data = salamanders_and_wood_and_water_and_stream)

summary(sala.lmm3)
plot(sala.lmm3)

#max depth
sala.lmm4 <- lme(avg_length ~ SITE + REACH_TYPE + average_maximum_depth, random =~1 | REACH_TYPE/SITE, data = salamanders_and_wood_and_water_and_stream)

summary(sala.lmm4)
plot(sala.lmm4)

#canopy cover
sala.lmm5 <- lme(avg_length ~ SITE + REACH_TYPE + average_canopy_cover, random =~1 | REACH_TYPE/SITE, data = salamanders_and_wood_and_water_and_stream)

summary(sala.lmm5)
plot(sala.lmm5)

=======

### Model 3
```{r}
# 1. Wilcoxon Rank-Sum Test: OG vs SG 
wilcox_result <- wilcox.test(Abundance_Estimate ~ REACH_TYPE, data = full_salamander_data)
print(wilcox_result)

# Boxplot to visualize
ggplot(full_salamander_data, aes(x = REACH_TYPE, y = Abundance_Estimate, fill = REACH_TYPE)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("seagreen3", "dodgerblue")) +
  labs(
    title = "Salamander Abundance by Forest Type",
    x = "Forest Type",
    y = "Abundance"
  ) +
  theme_bw() +
  theme(legend.position = "none")



# 2. Spearman Correlation: Abundance vs DEPTH_MAX 
depth_corr <- cor.test(full_salamander_data$Abundance_Estimate,
                       full_salamander_data$DEPTH_MAX,
                       method = "spearman")
print(depth_corr)

# Scatterplot
ggplot(full_salamander_data, aes(x = DEPTH_MAX, y = Abundance_Estimate)) +
  geom_point(color = "darkgreen", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(
    title = "Abundance vs. Maximum Pool Depth",
    x = "Maximum Pool Depth (cm)",
    y = "Abundance"
  ) +
  theme_bw()



# 3. Spearman Correlation: Abundance vs WOOD_LENGTH
wood_corr <- cor.test(full_salamander_data$Abundance_Estimate,
                      full_salamander_data$WOOD_LENGTH,
                      method = "spearman")
print(wood_corr)

# Scatterplot
ggplot(full_salamander_data, aes(x = WOOD_LENGTH, y = Abundance_Estimate)) +
  geom_point(color = "sienna4", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(
    title = "Abundance vs. Woody Debris Length",
    x = "Woody Debris Length (cm)",
    y = "Abundance"
  ) +
  theme_bw()
>>>>>>> 5e9cf5b72a1a637f839dbdf5b09db434eda47959
```

